---
title: "Lab 07 — t-tests and ANOVA"
author: "EAES 480 — Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Overview

This lab extends hypothesis testing from **claims about one mean** to:

- **one-sample t-tests**
- **two-sample t-tests**
- **paired t-tests**
- **ANOVA**

You will use the **US-AMS simplified AmeriFlux dataset** again and compare means across different time-based groups.

The core logic stays the same:

> estimate → uncertainty → test statistic → p-value

This lab emphasizes both **manual understanding** and **automated testing in R**.

---

# Learning goals

By the end of this lab, you should be able to:

- Compute and interpret a **one-sample t-test**
- Distinguish between **two-sample** and **paired** t-tests
- Compute and interpret **degrees of freedom**
- Explain the structure of an **ANOVA**
- Compare manual calculations to `t.test()` and `aov()`
- Use visualizations to support statistical interpretation
- Produce a fully reproducible R Markdown analysis

---

# Data

## Load and inspect

```{r load_data, echo=TRUE, eval=TRUE}
# GOAL: Read the dataset and keep 2022–2023.

df <- read_csv("data/us-ams-simple.csv", na = c("-9999")) %>%
  clean_names() %>%
  filter(year_local %in% c(2022, 2023))

glimpse(df)
```

## Derive time columns if needed

```{r derive_time, echo=TRUE, eval=TRUE}
# GOAL: Create date / month columns if they do not already exist.

df <- df %>%
  mutate(
    date = as.Date(floor(doy) - 1, origin = paste0(year_local, "-01-01")),
    month = month(date, label = TRUE, abbr = TRUE)
  )
glimpse(df)
```

---

# Part I — One-sample t-test

## 1) Choose a day and a variable

Select:
- one **year**
- one **numeric variable**

Examples:
- `ta`
- `swc`
- `le`
- `fc`

```{r choose_one_sample, echo=TRUE, eval=TRUE}
set.seed(480)
target_year <- 2023
target_var  <- 'ta'

sample_data <- df %>%
  filter(year_local == target_year) %>% 
  slice_sample(n = 100)

x <- sample_data[[target_var]]

mean(x, na.rm = TRUE); sd(x, na.rm = TRUE); length(x)
```

> What variable and day did you choose, and why?

---

>  - I chose ta (air temperature) and randomly sampled 100 observations from the year 2023. I chose this variable because temperature is easy to interpret and is expected to vary over time, making it a good candidate for hypothesis testing.

---

## 2) State a null hypothesis

Write a claim about the mean of your chosen variable on that day.

Examples:
- “The mean temperature on this day is 15 °C.”
- “The mean latent heat flux on this day is 100 W m$^{-2}$.”

Fill in:

> $H_0$: `The mean air temperature in my 2023 sample is 14.156 °C`

> $H_A$: `The mean air temperature in my 2023 sample is not 14.156 °C`

---

## 3) Compute the one-sample t-statistic manually

Use:

$t = (\bar{x} - \mu_0) / (s / \sqrt{n})$

```{r one_sample_manual, echo=TRUE, eval=TRUE}

mu0 <- 14.156

xbar <- mean(x, na.rm = TRUE)
s <- sd(x, na.rm = TRUE)
n <- sum(!is.na(x))

t_stat <- (xbar - mu0) / (s / sqrt(n))
t_stat
```

---

## 4) Compute the p-value and degrees of freedom

```{r one_sample_pvalue, echo=TRUE, eval=TRUE}

df_t <- n-1

p_value <- 2 * pt(-abs(t_stat), df = df_t)

tibble(
  xbar = xbar,
  s = s,
  n = n,
  df_t = df_t,
  t_stat = t_stat,
  p_value = p_value
)
```

> What does the p-value suggest about your null hypothesis?

---

> The p-value is 0.814, which is relatively large. Therefore, I would not reject the null hypothesis at the 0.05 level. My sample does not provide strong evidence that the mean air temperature differs from 14.156 °C.

---

## 5) Compare with `t.test()`

```{r one_sample_ttest, echo=TRUE, eval=TRUE}
t.test(x, mu = mu0)
```

> Does the `t.test()` output agree with your manual calculation? If not, why might that be? 

---

>  Yes, the t.test() output agrees with my manual calculation. The t-statistic, degrees of freedom, and p-value are essentially the same, with only tiny differences due to rounding.

---

# Part II — Two-sample t-test

## 6) Compare the same variable across two day ranges

Choose:
- the **same variable**
- two different **day ranges in 2023**

```{r choose_two_sample, echo=TRUE, eval=TRUE}

day_range1 <- "90_120"
day_range2 <- "300_330"
x1 <- df %>%
  filter(year_local == 2023, doy > 90 & doy < 120) %>%
  pull(target_var)

x2 <- df %>%
  filter(year_local == 2023, doy > 300 & doy < 330) %>%
  pull(target_var)

mean(x1, na.rm = TRUE); sd(x1, na.rm = TRUE); length(x1)
mean(x2, na.rm = TRUE); sd(x2, na.rm = TRUE); length(x2)
```

> Why did you choose these two days?

---

> I chose these two day ranges because they represent different parts of the year, and I expected air temperature to differ seasonally between them.

---

## 7) Summarize the difference in means

```{r two_sample_summary, echo=TRUE, eval=TRUE}
xbar1 <- mean(x1, na.rm = TRUE)
xbar2 <- mean(x2, na.rm = TRUE)

s1 <- sd(x1, na.rm = TRUE)
s2 <- sd(x2, na.rm = TRUE)

n1 <- sum(!is.na(x1))
n2 <- sum(!is.na(x2))

mean_diff <- xbar1 - xbar2

tibble(
  xbar1 = xbar1,
  xbar2 = xbar2,
  mean_diff = mean_diff
)
```

---

## 8) Compute the two-sample t-statistic manually (Welch)

```{r two_sample_manual, echo=TRUE, eval=TRUE}
se_diff <- sqrt((s1^2 / n1) + (s2^2 / n2))

t_stat <- mean_diff / se_diff

df_welch <- ((s1^2 / n1 + s2^2 / n2)^2) /
  (((s1^2 / n1)^2 / (n1 - 1)) + ((s2^2 / n2)^2 / (n2 - 1)))

p_value <- 2 * pt(-abs(t_stat), df = df_welch)

tibble(
  mean_diff = mean_diff,
  se_diff = se_diff,
  t_stat = t_stat,
  df_welch = df_welch,
  p_value = p_value
)
```

> What does the sign of `mean_diff` mean? What does the p-value mean? - Mean_diff = 4.45. The sign of mean_diff is positive, which means the first group (DOY 90–120) had the higher mean air temperature. The p-value is extremely small (1.20e-68), so I would reject the null hypothesis of equal means. This suggests very strong evidence that the average air temperature differed between these two day ranges in 2023.

---

## 9) Compare with `t.test()`

```{r two_sample_ttest, echo=TRUE, eval=TRUE}
t.test(x1, x2)
```

> Does the automated t-test agree with your manual result? - Yes, the automated two-sample t-test agrees with my manual result. Both approaches indicate a large positive difference in means and an extremely small p-value, so the conclusion is the same: the two groups differ strongly in mean air temperature.

---

## 10) Visualize the two groups

```{r two_sample_plot, echo=TRUE, eval=TRUE}
plot_df <- bind_rows(
  tibble(group = paste0("DOY_", day_range1), value = x1),
  tibble(group = paste0("DOY_", day_range2), value = x2)
)

ggplot(plot_df, aes(x = group, y = value, fill = group)) +
  geom_boxplot(alpha = 0.8) +
  theme_classic(base_size = 18) +
  labs(
    x = NULL,
    y = target_var,
    title = "Two-sample comparison"
  )
```

> Does the plot match your statistical conclusion? - Yes, the plot matches my statistical conclusion. The boxplots show that the DOY 90–120 group has generally higher air temperatures than the DOY 300–330 group, which is consistent with the positive mean difference and the very small p-value.

---

# Part III — Paired t-test

## 11) Compare the same days across two years

Choose one DOY and compare **2022 vs 2023** for the same variable.

```{r paired_setup, echo=TRUE, eval=TRUE}
x1 <- df %>%
  filter(year_local == 2022, doy > 180 & doy < 210) %>%
  arrange(date) %>%
  pull(target_var)

x2 <- df %>%
  filter(year_local == 2023, doy > 180 & doy < 210) %>%
  arrange(date) %>%
  pull(target_var)

length(x1); length(x2)
```

> Why is this a paired comparison rather than an independent two-sample comparison? - This is a paired comparison because the observations come from the same period of days in two different years, so each value in 2022 is naturally matched with a corresponding value in 2023. The data are not being treated as two unrelated groups; instead, the test focuses on the within-pair differences.

---

## 12) Compute paired differences manually

```{r paired_manual, echo=TRUE, eval=TRUE}
d <- x1 - x2

dbar <- mean(d, na.rm = TRUE)
sd_d <- sd(d, na.rm = TRUE)
n <- sum(!is.na(d))

t_stat <- dbar / (sd_d / sqrt(n))
p_value <- 2 * pt(-abs(t_stat), df = n - 1)

tibble(
  dbar = dbar,
  sd_d = sd_d,
  n = n,
  t_stat = t_stat,
  p_value = p_value
)
```

> What does the sign of `dbar` mean? 

> - Because I defined the differences as x1 - x2, the positive value of dbar means that air temperature was higher on average in 2022 than in 2023 for this same day-of-year range

---

## 13) Compare with `t.test(..., paired = TRUE)`

```{r paired_ttest, echo=TRUE, eval=TRUE}
t.test(x1, x2, paired = TRUE)
```

> How does the paired t-test differ conceptually from the two-sample t-test?

---

> The paired t-test differs from the two-sample t-test because it uses matched observations and tests whether the mean of the within-pair differences is zero. In contrast, a two-sample t-test treats the two groups as independent. The paired test is appropriate here because each observation in one year is matched to the same seasonal window in the other year.

---

# Part IV — ANOVA

## 14) Create a categorical grouping variable

Group `ta` into broad seasonal categories:

- Summer
- Winter
- Other

```{r make_groups_anova, echo=TRUE, eval=TRUE}
df_anova <- df %>%
  filter(!is.na(ta)) %>%
  mutate(
    month_group = case_when(
      month %in% c('Jun', 'Jul', 'Aug') ~ "Summer",
      month %in% c('Dec', 'Jan', 'Feb') ~ "Winter",
      TRUE ~ "Other"
    )
  )

df_anova %>% count(month_group)
```

---

## 15) Visualize the groups

```{r plot_anova_groups, echo=TRUE, eval=TRUE}
ggplot(df_anova, aes(x = month_group, y = ta, fill = month_group)) +
  geom_boxplot(alpha = 0.8) +
  theme_classic(base_size = 18) +
  labs(
    x = NULL,
    y = "ta",
    title = "Air temperature by seasonal group"
  )
```

> Do the groups look different enough that ANOVA might detect a difference?

---

>
Yes, the groups look different enough that ANOVA is very likely to detect a difference. In particular, the Summer, Winter, and Other categories show clear separation in air temperature, so I would expect a significant result.

---

## 16) Compute ANOVA components manually

```{r anova_manual_components, echo=TRUE, eval=TRUE}
grand_mean <- mean(df_anova$ta, na.rm = TRUE)

group_stats <- df_anova %>%
  group_by(month_group) %>%
  summarise(
    group_mean = mean(ta, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

ss_between <- sum(group_stats$n * (group_stats$group_mean - grand_mean)^2)

df_anova2 <- df_anova %>%
  left_join(group_stats, by = "month_group")

ss_within <- sum((df_anova2$ta - df_anova2$group_mean)^2)

k <- n_distinct(df_anova$month_group)
n_total <- nrow(df_anova)

df_between <- k - 1
df_within <- n_total - k

ms_between <- ss_between / df_between
ms_within <- ss_within / df_within

f_stat <- ms_between / ms_within
p_value <- pf(f_stat, df1 = df_between, df2 = df_within, lower.tail = FALSE)

tibble(
  grand_mean = grand_mean,
  ss_between = ss_between,
  ss_within = ss_within,
  df_between = df_between,
  df_within = df_within,
  f_stat = f_stat,
  p_value = p_value
)
```

> In your own words, what do `ss_between` and `ss_within` represent?

---

>ss_between represents the variation explained by the group structure, meaning how much the group means differ from the grand mean. ss_within represents the variation that remains inside each group, meaning how much individual observations differ from their own group mean. 

---


## 17) Compare with `aov()`

```{r run_anova, echo=TRUE, eval=TRUE}
model <- aov(ta ~ month_group, data = df_anova)
summary(model)
```

> What does the ANOVA p-value tell you? Does it tell you which groups differ?

---

>
The ANOVA p-value is essentially zero (<2e-16), so I would reject the null hypothesis that all group means are equal. This tells me that at least one seasonal group has a different mean air temperature. However, ANOVA by itself does not tell me exactly which groups differ from each other; I would need post-hoc comparisons to determine that.

---

## 18) Final reflection

Answer in **5–7 sentences**:

- How is a one-sample t-test different from a two-sample t-test?
- Why is a paired t-test special?
- How is ANOVA a generalization of the t-test?
- What stayed the same across all these tests?

> A one-sample t-test compares a sample mean to a specific value, while a two-sample t-test compares two groups. A paired t-test is different because it uses matched observations and focuses on the differences between pairs. This works well when the data are naturally linked, like the same days in two different years. ANOVA is similar but compares more than two groups at once. In all of these tests, the main idea is the same: estimate a difference, consider uncertainty, calculate a test statistic, and use a p-value to decide whether to reject the null hypothesis. So even though the methods look different, the underlying idea is very similar.

---

# Submission

- Knit your `.Rmd` to HTML.
- Commit and push both the `.Rmd` and knitted `.html` to your GitHub repo.
- Submit the GitHub repository link on Canvas.
